Do I Need to Model You in order to Learn With You?

### 2015-04-01 mutual-modeling.tex
Fixed a few ref + style + formatting in the first paragraphs of article

From its very beginning, CSCL research has been following
\citet{roschelle1995construction} suggestion that collaborative learning has
something to do with the process of constructing and maintaining a \emph{shared
understanding} of the task at hand. Building a shared/mutual understanding
refers to the upper class of collaborative learning situations, those in which
students should build upon each other's understanding to refine their own
understanding.  What is expected to produce learning is not the mere fact that
two students build the same understanding but the cognitive effort they have to
engage to build this shared understanding~\citep{schwartz1995emergence}. This
effort can be observed by the frequency of rich interactions, \ie interactions
whose occurrence has been related to learning: (self-) explanations in cognitive
science (Chi ; Webb), conflict resolution in socio-cognitive theories (Doise),
mutual regulation (Blaye,) in a Vygostkian perspective, etc. The construction of
a shared understanding has been investigated for several years in
psycholinguistics, under the  notion of
``grounding''~\citep{clark1986referring}.  However, the relevance of grounding
mechanisms for explaining learning outcomes has been questioned in learning
sciences. The monitoring and repair of misunderstanding explains for instance
referential failures in short dialogue episodes but does hardly predict
\emph{conceptual change} (\ie the acquisition, acceptation and integration of a
new belief into one's mental model\fxnote{check with Pierre}) over longer
sessions~\cite{dillenbourg2006sharing}. The cumulative effect of grounding
episodes can probably be better understood from a socio-cultural perspective:

@@

Along this line, several scholars suggest that CSCL research should go deeper in
the understanding of how partners engage into shared meaning
making~\cite{stahl2007meaning} or \emph{intersubjective} meaning
making~\cite{suthers2006technology}.

Paradoxically, while Clark's theory is somewhat too linguistic from a conceptual
change viewpoint, it is criticized at the same time as being too cognitivist by
some psycholinguists, \ie as overestimating the amount of shared knowledge and
mutual representations actually necessary to conduct a dialogue. The fundamental
issue, as old as philosophy, is the degree of coupling between the different
levels of dialogue, mostly between the lexical/syntactical level and the deeper
semantic levels. \citet{pickering2006alignment} argue that the mutual
understanding starts mostly with a \emph{superficial alignment} at the level of
the linguistic representations, due to priming mechanisms, and that this local
alignment may -- in some cases -- lead to a \emph{global alignment} of the
semantic level (\emph{deep grounding}).  For these authors, the convergence in
dialogue, and even the repair of some misunderstandings, is explained by this
mimetic behavior more than by a monitoring of each other's knowledge:
\emph{``...interlocutors do not need to monitor and develop full common ground
as a regular, constant part of routine conversation, as it would be
unnecessary and far too costly. Establishment of full common ground is, we
argue, a specialized and non-automatic process that is used primarily in
times of difficulty (when radical misalignment becomes apparent).''} (p. 179).
This view is actually not incompatible with Clark's \emph{grounding
criterion}~\cite{clark1989contributing}: the degree of shared understanding that
peers need to reach depends upon the task they perform. For instance, a dialogue
between two surgeons might rely on superficial alignment if they talk about
their friends but has to guarantee accurate common grounds when talking about
which intervention will be conducted in which way on which patient.  Some CSCL
scholars stressed the ``illusion of shared understanding'' \fixme{REF:(JACCO?)},
\ie the potential decoupling between linguistic alignment and actual shared
meanings.

This interesting cognitive science debate mostly occurred outside the field of
learning. In education, the question is to relate these mechanisms to learning
outcomes. Is linguistic alignment sufficient to trigger conceptual change? Does
negotiation of meaning only occurs when partners monitor and diagnose each
other's knowledge. If the ratio between shallow alignment and deep grounding
depends upon the task, and if deep grounding is a condition for learning, then
the pedagogical challenge is to design tasks that require deep grounding. Most
empirical studies on grounding and alignment are conducted with tasks, which
despite being qualified as ``ecologically valid'' by their authors, are mere
referencing tasks such as asking the way to the train station or helping the
peer to choose a picture among many. In this contribution, we explore several
richer tasks such as arguing on a sensitive issue or building a concept map.

Deep grounding or shared meaning making requires some cognitive load. For
\citet{clark1986referring}, what is important is not the individual effort made
by the receiver of a communicative act, but the overall \emph{least collaborative
effort}.  The cost of producing a perfect utterance may be higher than the cost
of repairing the problems that may arise through misunderstandings. For
instance, subjects are less careful about adapting their utterances to their
partner when they know they can provide feedback on his/her
understanding~\citep{schober1993spatial}. We introduced the notion of
\emph{optimal collaborative effort}~\citep{dillenbourg1995evolution} to stress
that misunderstanding should not be viewed as something to be avoided (if this
was possible), but as an opportunity to engage into verbalization, explanation,
negotiation, and so forth. This issue is related to the global argument
regarding cognitive load in learning activities, especially in discovery
learning environments: there is no learning without some cognitive load, but
overload may hinder learning~\citep{paas2003cognitive}. In the context of
collaborative learning, we understand the cognitive load induced by mutual
modelling as part of~\citet{schwartz1995emergence} notion of effort towards a
shared understanding. For instance, CSCL researchers expanded the use of
\emph{collaboration scripts} \fixme{REF (Aronson,?)}. A script is a pedagogical
method that collaborative learning activities in order to foster the emergence
of productive interactions such as argumentation, explanation or conflict.
Conflict-resolution scripts such as the {\sc
ArgueGraph}~\citep{dillenbourg2008mechanics} forms pair of students with
opposite opinions, which increases the difficulty of consensus building,
requiring more justifications, more negotiation, more load. Similarly, JIGSAW
scripts provide peers with different but complementary knowledge for augmenting
(reasonably) the effort group members have to engage to reach a shared solution.

%\paragraph{Theory of Mind}
%
%Theory of Mind (originally defined in~\citep{premack1978does}) is the cognitive
%ability that a subject possesses to represent the mental state of another
%agent, possibly including knowledge that contradicts the subject's own model: for
%example, a book can be at the same time \emph{visible} for myself, and \emph{not
%visible} for you.
%
%Children develop this skill, which is essential to understand others'
%perspectives during interactions, around the age of three. It supposes the
%ability to build, store and retrieve separate models of the knowledge of the
%interactors.
%
%One classical application of this cognitive skill is the so-called
%\emph{False-Belief} experiment (also known as the \emph{Sally and Ann}
%experiment)~\citep{Leslie2000}: a child is asked to watch a scene where two
%people, A and B, manipulate objects. Then A leaves and B hides away one
%object. When A comes back, we ask the child ``where do you think A will
%look for the object?''. Before acquiring a theory of mind, children are not
%able to separate their own (true) model of the world (where they know that
%the object was hidden) from the model of A, which contains \emph{false
%beliefs} on the world (A still thinks the object is at its original
%position since he did not see B hiding it).

@@

In order to build a shared understanding, do partners have to build a
representation of each other's knowledge? We refer to \emph{mutual modelling} as
the process of inferring one's partner mental states. Any claim that students
carry out a detailed monitoring of their peers would be as incorrect as any
claim that they do not maintain any representation at all. If mutual modelling
had to be permanently detailed and accurate, subjects would obviously face a
huge cognitive load. Conversely, peers could not collaborate without some
minimal amount of mutual modelling. For instance, A cannot disagree with B
without knowing that B has a different opinion. The mutual model can be implicit
(A is not aware of what he knows about B), by-default (I believe that B beliefs
what I believe unless contrary evidence), opportunistic (A does not model B
unless the conversation requires it), global (A infers B's beliefs based on
categories such as age, culture or profession) and, of course, it can be
incorrect… but it can not remain empty. Dialogues include many instances of
utterances such as ``I thought he would do that'' (first level of mutual
modelling) or even ``He thought I would do that but I intended something else.''
(second level of mutual modelling).

The content of mutual models ranges from 'dispositional' versus 'situational'
aspects. The 'dispositional' aspects refer to A's representation of B's long
term knowledge, skills or traits. It is thus closely related to the notion of
transactive memory~\citep{wegner1987transactive, moreland1999transactive}.
'Situational' aspects refer to A's representation of B's knowledge, behavior or
intentions specifically activated in the situation in which A and B are
collaborating, some of them being valid for 2 seconds, other ones for 2 hours.
Here are examples of fragments that constitute A's model of B regarding to
aspects X, noted $Model(A,B,X)$:

@@

    \item $Model(A,B, knowledge)$: What does A know about B's knowledge with
        respect to the task at hand or, inversely, about B's knowledge gaps?
        When can A consider B's statements as reliable?

    \item $Model(A,B, skills)$: What does A know about B's skills with respect to
        the task at hand? May A expect B to perform well in a specific subtask?
        The effectiveness of division of labor depends on the quality of this
        mutual model.

    \item $Model(A,B, goals)$: What does A know about B's intentions with respect
        to the project, including B' motivation and commitment? Can A trust B
        when B promises to deliver?

    \item $Model(A,B, task)$: What does A know about B's representation of the
        situation and the task: does A knows whether B has the same
        understanding of the problem at stake?

    \item $Model(A,B, plans)$: What does A know about B's strategy. Does A
        understand why B did what he did? Is A able to anticipate what B will do
        next?

    \item $Model(A,B, ``urgent'')$: What does A about know B's understanding of A's
        last utterance: does ‘urgent' means now, ASAP and ‘not too lat' ?

    \item We could continue the list of what is X is \model{A}{B}{X}: beliefs,
        emotions, history, status, …

@@

We have different levels of mutual modelling. If A states ``B thinks I am good in
maths'', we are the second level of mutual modelling:  $Model(A, B, Model(B,A,
knowledge=``good in math''))$. There is an infinite regress of nested models: If A
says ``B knows that I don't expect him to solve this statistics problem''
corresponds to $Model(A, B, Model (B, A (Model (A,B, statistic-skills)))$.

>>>>

From its very beginning, CSCL research has been following
\citet{roschelle1995construction} suggestion that collaborative learning has
something to do with the process of constructing and maintaining a \emph{shared
understanding} of the task at hand. Building a shared/mutual understanding
refers to the upper class of collaborative learning situations, those in which
students should build upon each other's understanding to refine their own
understanding: what is expected to produce learning is not the mere fact that
two students build the same understanding but the cognitive effort they have to
engage to build this shared understanding~\citep{schwartz1995emergence}. This
effort can be observed by the frequency of rich interactions, \ie interactions
whose occurrence has been related to learning: (self-) explanations in cognitive
science \fixme{REF:Chi} \fixme{REF: Webb}, conflict resolution in
socio-cognitive theories \fixme{REF:Doise},
mutual regulation \fixme{REF: Blaye} in a Vygostkian perspective, etc. The
construction of a shared understanding has been investigated for several years
in psycholinguistics, under the  notion of
\emph{grounding}~\citep{clark1986referring}. However, the relevance of grounding
mechanisms for explaining learning outcomes has been questioned in learning
sciences. The monitoring and repair of misunderstanding explains for instance
referential failures in short dialogue episodes but does hardly predict
\emph{conceptual change} (\ie the acquisition, acceptation and integration of a
new belief into one's mental model\fxnote{check with Pierre}) over longer
sessions~\cite{dillenbourg2006sharing}. The cumulative effect of grounding
episodes can probably be better understood from a socio-cultural perspective:

@@

Along this line, several scholars suggest that CSCL research should go deeper in
the understanding of how partners engage into shared meaning
making~\citep{stahl2007meaning} or \emph{intersubjective} meaning
making~\citep{suthers2006technology}.

Paradoxically, while Clark's theory is somewhat too linguistic from a conceptual
change viewpoint, it is criticized at the same time as being too cognitivist by
some psycholinguists, \ie as overestimating the amount of shared knowledge and
mutual representations actually necessary to conduct a dialogue. The fundamental
issue, as old as philosophy, is the degree of coupling between the different
levels of dialogue, mostly between the lexical/syntactical level and the deeper
semantic levels. \citet{pickering2006alignment} argue that the mutual
understanding starts mostly with a \emph{superficial alignment} at the level of
the linguistic representations, due to priming mechanisms, and that this local
alignment may -- in some cases -- lead to a \emph{global alignment} of the
semantic level (\emph{deep grounding}).  For these authors, the convergence in
dialogue, and even the repair of some misunderstandings, is explained by this
mimetic behavior more than by a monitoring of each other's knowledge:
\emph{``...interlocutors do not need to monitor and develop full common ground
as a regular, constant part of routine conversation, as it would be
unnecessary and far too costly. Establishment of full common ground is, we
argue, a specialized and non-automatic process that is used primarily in
times of difficulty (when radical misalignment becomes apparent).''} (p. 179).
This view is actually not incompatible with Clark's \emph{grounding
criterion}~\citep{clark1989contributing}: the degree of shared understanding that
peers need to reach depends upon the task they perform. For instance, a dialogue
between two surgeons might rely on superficial alignment if they talk about
their friends but has to guarantee accurate common grounds when talking about
which intervention will be conducted in which way on which patient. 
%Some CSCL
%scholars stressed the ``illusion of shared understanding'' \fixme{REF: Jacco?},
%\ie the potential decoupling between linguistic alignment and actual shared
%meanings.

This interesting cognitive science debate occurred mostly outside the field of
learning. In education, the question is to relate these mechanisms to learning
outcomes: \textbf{Is linguistic alignment sufficient to trigger conceptual
change?} \textbf{Does negotiation of meaning only occurs when partners monitor
and diagnose each other's knowledge?} If the ratio between shallow alignment and
deep grounding depends upon the task, and if deep grounding is a condition for
learning, then the pedagogical challenge is indeed to design tasks that require
deep grounding. Most empirical studies on grounding and alignment are conducted
with tasks, which despite being qualified as ``ecologically valid'' by their
authors, are mere referencing tasks such as asking the way to the train station
or helping the peer to choose a picture among many. In this contribution, we
explore several richer tasks such as arguing on a sensitive issue or building a
concept map.

Deep grounding or shared meaning making requires some cognitive load. For
\citet{clark1986referring}, what is important is not the individual effort made
by the receiver of a communicative act, but the overall \emph{least collaborative
effort}.  The cost of producing a perfect utterance may be higher than the cost
of repairing the problems that may arise through misunderstandings, and in fact,
subjects tend to make less efforts adapting their utterances to a specific
partner when they know that they can later provide feedback on his/her
understanding~\citep{schober1993spatial}. We introduced the notion of
\emph{optimal collaborative effort}~\citep{dillenbourg1995evolution} to stress
that misunderstanding should not be viewed as something to be avoided (if this
was possible), but as an opportunity to engage into verbalization, explanation,
negotiation, and so forth. This issue is related to the global argument
regarding cognitive load in learning activities, especially in discovery
learning environments: there is no learning without some cognitive load, but
overload may hinder learning~\citep{paas2003cognitive}. In the context of
collaborative learning, we understand the cognitive load induced by mutual
modelling as part of~\citet{schwartz1995emergence} notion of effort towards a
shared understanding. For instance, CSCL researchers expanded the use of
\emph{collaboration scripts}~\citep{kobbe2007specifying}. A script is a pedagogical
method that frames collaborative learning activities in order to foster the
emergence of productive interactions such as argumentation, explanation or
conflict. Conflict-resolution scripts such as the {\sc
ArgueGraph}~\citep{dillenbourg2008mechanics} form pair of students with
opposite opinions, which increases the difficulty of consensus building,
requiring more justifications, more negotiation, more load. Similarly, {\sc
jigsaw} scripts~\citep{aronson1978jigsaw} provide peers with different but
complementary knowledge for augmenting (reasonably) the efforts that group members
have to engage into to reach a shared solution.

%\paragraph{Theory of Mind}
%
%Theory of Mind (originally defined in~\citep{premack1978does}) is the cognitive
%ability that a subject possesses to represent the mental state of another
%agent, possibly including knowledge that contradicts the subject's own model: for
%example, a book can be at the same time \emph{visible} for myself, and \emph{not
%visible} for you.
%
%Children develop this skill, which is essential to understand others'
%perspectives during interactions, around the age of three. It supposes the
%ability to build, store and retrieve separate models of the knowledge of the
%interactors.
%
%One classical application of this cognitive skill is the so-called
%\emph{False-Belief} experiment (also known as the \emph{Sally and Ann}
%experiment)~\citep{Leslie2000}: a child is asked to watch a scene where two
%people, $A$ and $B$, manipulate objects. Then $A$ leaves and $B$ hides away one
%object. When $A$ comes back, we ask the child ``where do you think $A$ will
%look for the object?''. Before acquiring a theory of mind, children are not
%able to separate their own (true) model of the world (where they know that
%the object was hidden) from the model of A, which contains \emph{false
%beliefs} on the world ($A$ still thinks the object is at its original
%position since he did not see $B$ hiding it).

\vspace{2em}
These preliminary remarks frame the main question that this article address:
in order to engage into shared understanding, which is a desirable learning
situation, do partners have to build a representation of each other's knowledge?
if so, what exactly is the nature of this mutual modelling? We adopt hereafter a
constructive approach, first introducing a simple yet convenient model of mutual
modelling, then analyzing five independent studies in terms of the interactions
between learning and mutual modelling.

@@

We refer to \emph{mutual modelling} as the process of inferring one's partner
mental states. Any claim that students carry out a detailed monitoring of their
peers would be as incorrect as any claim that they do not maintain any
representation at all. If mutual modelling had to be permanently detailed and
accurate, subjects would obviously face a huge cognitive load. Conversely, peers
could not collaborate without some minimal amount of mutual modelling. For
instance, $A$ cannot disagree with $B$ without knowing that $B$ has a different
opinion. The mutual model can be implicit ($A$ is not aware of what he knows about
B), by-default (I believe that $B$ beliefs what I believe unless contrary
evidence), opportunistic ($A$ does not model $B$ unless the conversation requires
it), global ($A$ infers $B$'s beliefs based on categories such as age, culture or
profession) and, of course, it can be incorrect… but it can not remain empty.
Dialogues include many instances of utterances such as ``I thought he would do
that'' (first level of mutual modelling) or even ``He thought I would do that
but I intended something else.'' (second level of mutual modelling).

The content of mutual models ranges from \emph{dispositional} versus
\emph{situational} aspects. The \emph{dispositional} aspects refer to $A$'s
representation of $B$'s long term knowledge, skills or traits. It is thus closely
related to the notion of transactive memory~\citep{wegner1987transactive,
moreland1999transactive}.  \emph{Situational} aspects refer to $A$'s representation of
B's knowledge, behavior or intentions specifically activated in the situation in
which $A$ and $B$ are collaborating, some of them being valid for 2 seconds, other
ones for 2 hours.  Here are examples of fragments that constitute $A$'s model of B
regarding to aspects X, noted $Model(A,B,X)$:

@@

    \item $Model(A,B, knowledge)$: What does $A$ know about $B$'s knowledge with
        respect to the task at hand or, inversely, about $B$'s knowledge gaps?
        When can $A$ consider $B$'s statements as reliable?

    \item $Model(A,B, skills)$: What does $A$ know about $B$'s skills with respect to
        the task at hand? May $A$ expect $B$ to perform well in a specific subtask?
        The effectiveness of division of labor depends on the quality of this
        mutual model.

    \item $Model(A,B, goals)$: What does $A$ know about $B$'s intentions with respect
        to the project, including $B$' motivation and commitment? Can $A$ trust B
        when $B$ promises to deliver?

    \item $Model(A,B, task)$: What does $A$ know about $B$'s representation of the
        situation and the task: does $A$ knows whether $B$ has the same
        understanding of the problem at stake?

    \item $Model(A,B, plans)$: What does $A$ know about $B$'s strategy. Does A
        understand why $B$ did what he did? Is $A$ able to anticipate what $B$ will do
        next?

    \item $Model(A,B, ``urgent'')$: What does $A$ about know $B$'s understanding of $A$'s
        last utterance: does ‘urgent' means now, ASAP and ‘not too lat' ?

    \item We could continue the list of what is $X$ is \model{A}{B}{X}: beliefs,
        emotions, history, status, …

@@

We have different levels of mutual modelling. If $A$ states ``B thinks I am good in
maths'', we are the second level of mutual modelling:  $Model(A, $B$, Model(B,A,
knowledge=``good in math''))$. There is an infinite regress of nested models: If A
says ``B knows that I don't expect him to solve this statistics problem''
corresponds to $Model(A, $B$, Model (B, A (Model (A,B, statistic-skills)))$.
